\Large Feature and order manipulations in a free recall task affect memory for current and future lists

### 2023-07-04 paper/main.tex
re-ran all notebooks, made some minor format changes per editor request
starting to address reviewer comments

Over the past half-century, context-based models have had impressive success at
explaining many stereotyped behaviors observed during free recall and other
list-learning tasks~\citep{Este55a, RaaiShif80, GlenEtal83, HowaKaha02a,
SiroEtal05, KimbEtal07, PolyKaha08, SedeEtal08, PolyEtal09, ShanHowa12}. These
phenomena include the well known recency and primacy effects (superior recall
of items from the end and, to a lesser extent, from the beginning of the study
list), as well as semantic and temporal clustering effects~\citep{KahaEtal08b,
HowaKaha02b}. The contiguity effect is an example of temporal clustering, which
is perhaps the dominant form of organization in free recall. This effect can be
seen in people's tendencies to successively recall items that occupied
neighboring positions in the studied list~\citep{Kaha96}. There are also
striking effects of semantic clustering~\citep{RomnEtal93, Bous53, BousEtal54,
JenkRuss52, MannKaha12}, whereby the recall of a given item is more likely to
be followed by recall of a similar or related item than a dissimilar or
unrelated one. In general, people organize memories for words along a wide
variety of stimulus dimensions. As formalized by models like the
\textit{Context Maintenance and Retrieval Model}~\citep{PolyEtal09}, the
stimulus features associated with each word (e.g.\ the word's meaning, size of
the object the word represents, the letters that make up the word, font size,
font color, location on the screen, etc.) are incorporated into the
participant's mental context representation~\citep{SmitVela01, MannEtal11,
MannEtal12, MannEtal15, Mann20}. During a memory test, any of these features
may serve as a memory cue, which in turn leads the participant to recall in
succession words that share stimulus features.

@@

\paragraph{Defining feature-based distances.} Sorting words according to a
given relevant feature requires first defining a distance function for
quantifying the dissimilarity between each pair of features. This function
varied according to the type of feature under consideration. Semantic features
(category and size) are \textit{categorical}. For these features, we defined a
binary distance function: two words were considered to ``match'' (i.e., have a
distance of 0) if their labels were the same (i.e., both from the same semantic
category or both of the same size). If two words' labels were different for a
given feature, we defined the words to have a distance of 1 for that feature.
Lexicographic features (length and first letter) are \textit{discrete}. For
these features we defined a discrete distance function. Specifically, we
defined the distance between two words as either the absolute difference
between their lengths, or the absolute distance between their starting letters
in the English alphabet, respectively. For example, two words that started with
the same letter would have a ``first letter'' distance of 0, and a pair of
words starting with `J' and `A' would have a first letter distance of 9.
Because words' lengths and letters' positions in the alphabet are always
integers, these discrete distances always take on integer values. Finally, the
visual features (color and location) are \textit{continuous} and
\textit{multivariate}, in that each ``feature'' is defined by multiple
(positive) real values. We defined the ``color'' and ``location'' distances
between two words as the Euclidean distances between their $(r, g, b)$ color or
$(x, y)$ location vectors, respectively. Therefore, the color and location
distance measures always take on non-negative real values (upper-bounded at
441.67 for color, or 27~in for location, reflecting the distances between the
corresponding maximally different vectors).

@@

The lag-conditional response probability (lag-CRP) curve~\citep{Kaha96}
reflects the probability of recalling a given item after the just-recalled
item, as a function of their relative encoding positions (lag). In other words,
a lag of 1 indicates that a recalled item was presented immediately after the
previously recalled item, and a lag of $-3$ indicates that a recalled item came
three items before the previously recalled item. For each recall transition
(following the first recall), we computed the lag between the just-recalled
word's presentation position and the next-recalled word's presentation
position. We computed the proportions of transitions (between successively
recalled words) for each lag, normalizing for the total numbers of possible
transitions. In carrying out this analysis, we excluded all incorrect recalls
and successive repetitions (i.e., recalling the same word twice in a row). This
yielded, for each list, a 1 by number-of-lags ($-15$ to $+15$; 30 lags in
total, excluding lags of 0) array of conditional probabilities. We averaged
these probabilities across lists to obtain a single lag-CRP for each
participant. Because transitions at large absolute lags are rare, these curves
are typically displayed using range restrictions~\citep{Kaha12}.

@@

@@

With these differences in mind, we next compared participants' memories on
early versus late lists for two additional experimental conditions (see
\textit{Random conditions}, Fig.~\dynamicsRandom). In a \textit{reduced
(early)} condition, we held the visual features constant on early lists, but
allowed them to vary randomly on late lists. In a \textit{reduced (late)}
condition, we allowed the visual features to vary randomly on early lists, but
held them constant on late lists. Given our above findings that (a)
participants tended to remember more words and exhibit stronger clustering
effects on feature rich (versus reduced) lists, and (b) participants tended to
remember more words and exhibit stronger clustering effects on early (versus
late) lists, we expected these early versus late differences to be enhanced in
the reduced (early) condition and diminished in the reduced (late) condition.
However, to our surprise, participants in \textit{neither} condition exhibited
reliable early versus late differences in accuracy (reduced (early): $t(41) =
1.499, p = 0.141$; reduced (late): $t(40) = 1.462, p = 0.152$), temporal
clustering (reduced (early): $t(41) = 0.998, p = 0.324$; reduced (late): $t(40)
= 1.099, p = 0.278$), nor feature-based clustering (reduced (early), category:
$t(41) = 0.753, p = 0.456$; reduced (early), size: $t(41) = 0.721, p = 0.475$;
reduced (early), length: $t(41) = 0.493, p = 0.625$; reduced (early), first
letter: $t(41) = 0.780, p = 0.440$; reduced (late), category: $t(40) = -0.086,
p = 0.932$; reduced (late), size: $t(40) = 0.746, p = 0.460$; reduced (late),
length: $t(40) = 1.476, p = 0.148$; reduced (late), first letter: $t(40) =
0.966, p = 0.340$). We hypothesized that adding or removing the variability in
the visual features was acting as a sort of ``event boundary'' between early
and late lists. In prior work, we (and others) have found that memories formed
just after event boundaries can be enhanced~\citep[e.g., due to less contextual
interference between pre- and post-boundary items;][]{MannEtal16, PettEtal16,
GoldEtal17, FlorEtal17}.

@@

If participants show different sensitivities to order manipulations, how do
their behaviors carry over to later lists? We found that participants who
showed strong feature clustering on early lists often tended to show strong
feature clustering on late lists (Fig.~\ref{fig:clustering-carryover}A; overall
across participants and conditions: $r(179) = 0.592, p < 0.001$; non-visual
feature conditions: all $r$s $\geq 0.350$, all $p$s $\leq 0.058$; color: $r(29)
= -0.071, p = 0.704$; location: $r(28) = 0.032, p = 0.868$; across conditions:
$r(4) = 0.934, p = 0.006$). Although participants tended to show weaker feature
clustering on late lists (Fig.~\ref{fig:fingerprint-trajectories}) on
\textit{average}, the associations between early and late lists for individual
participants suggests that some influence of early order manipulations may
linger on late lists. We found that participants who exhibited larger carryover
in feature clustering (i.e., continued to show strong feature clustering on
late lists) for the semantic order manipulations (but not other manipulations)
also tended to show a larger improvement in recall
(Fig.~\ref{fig:clustering-carryover}B; overall: $r(179) = 0.378, p < 0.001$;
category: $r(28) = 0.419, p = 0.021$; size: $r(28) = 0.737, p < 0.001$;
non-semantic conditions: all $r$s $\leq 0.252$, all $p$s $\geq 0.179$; across
conditions: $r(4) = 0.773, p = 0.072$) on late lists, relative to early lists.
Participants who exhibited larger carryover in feature clustering also tended
to show stronger temporal clustering on late lists (relative to early lists)
for all but the category condition (Fig.~\ref{fig:clustering-carryover}C;
overall: $r(179) = 0.434, p < 0.001$; category: $r(28) = 0.229, p = 0.223$; all
non-category conditions: all $r$s $\geq 0.448$, all $p$s $\leq 0.012$; across
conditions: $r(4) = 0.598, p = 0.210$).

@@

All of the data analyzed in this manuscript, along with all of the code for
carrying out the analyses may be found at
https://github.com/ContextLab/FRFR-analyses. Code for running the non-adaptive
experimental conditions may be found at
https://github.com/Con\-text\-Lab/efficient-learning-code. Code for running the
adaptive experimental condition may be found at
https://github.com/ContextLab/adaptiveFR. We have also released an associated
Python toolbox for analyzing free recall data, which may be found at
https://cdl-quail.read\-the\-docs.io/\-en/\-latest/.

>>>>

Over the past half-century, context-based models have had impressive success at
explaining many stereotyped behaviors observed during free recall and other
list-learning tasks~\citep{Este55a, RaaiShif80, GlenEtal83, HowaKaha02a,
SiroEtal05, KimbEtal07, PolyKaha08, SedeEtal08, PolyEtal09, ShanHowa12}. These
phenomena include the well known recency and primacy effects (superior recall
of items from the end and, to a lesser extent, from the beginning of the study
list), as well as semantic and temporal clustering effects~\citep{KahaEtal08b,
HowaKaha02b}. The contiguity effect is an example of temporal clustering, which
is perhaps the dominant form of organization in free recall. This effect can be
seen in people's tendencies to successively recall items that occupied
neighboring positions in the studied list~\citep{Kaha96}. There are also
striking effects of semantic clustering~\citep{RomnEtal93, Bous53, BousEtal54,
JenkRuss52, MannKaha12}, whereby the recall of a given item is more likely to
be followed by recall of a similar or related item than a dissimilar or
unrelated one. In general, people organize memories for words along a wide
variety of stimulus dimensions. According to models like the
\textit{Context Maintenance and Retrieval Model}~\citep{PolyEtal09}, the
stimulus features associated with each word (e.g.\ the word's meaning, size of
the object the word represents, the letters that make up the word, font size,
font color, location on the screen, etc.) are incorporated into the
participant's mental context representation~\citep{SmitVela01, MannEtal11,
MannEtal12, MannEtal15, Mann20}. During a memory test, any of these features
may serve as a memory cue, which in turn leads the participant to recall in
succession words that share stimulus features.

@@

\paragraph{Defining feature-based distances.} Sorting words according to a
given relevant feature requires first defining a distance function for
quantifying the dissimilarity between each pair of features. This function
varied according to the type of feature under consideration. Semantic features
(category and size) are \textit{categorical}. For these features, we defined a
binary distance function: two words were considered to ``match'' (i.e., have a
distance of 0) if their labels were the same (i.e., both from the same semantic
category or both of the same size). If two words' labels were different for a
given feature, we defined the words to have a distance of 1 for that feature.
Lexicographic features (length and first letter) are \textit{discrete}. For
these features we defined a discrete distance function. Specifically, we
defined the distance between two words as either the absolute difference
between their lengths, or the absolute distance between their starting letters
in the English alphabet, respectively. For example, two words that started with
the same letter would have a ``first letter'' distance of 0, and a pair of
words starting with `J' and `A' would have a first letter distance of 9.
Because words' lengths and letters' positions in the alphabet are always
integers, these discrete distances always take on integer values. Finally, the
visual features (color and location) are \textit{continuous} and
\textit{multivariate}, in that each ``feature'' is defined by multiple
(positive) real values. We defined the ``color'' and ``location'' distances
between two words as the Euclidean distances between their $(r, g, b)$ color or
$(x, y)$ location vectors (specified in inches), respectively. Therefore, the
color and location distance measures always take on non-negative real values
(upper-bounded at 441.67 for color, or 27~in for location, reflecting the
distances between the corresponding maximally different vectors).

@@

The lag-conditional response probability (lag-CRP) curve~\citep{Kaha96}
reflects the probability of recalling a given item after the just-recalled
item, as a function of their relative encoding positions (lag). In other words,
a lag of 1 indicates that a recalled item was presented immediately after the
previously recalled item, and a lag of $-3$ indicates that a recalled item came
three items before the previously recalled item. For each recall transition
(following the first recall), we computed the lag between the just-recalled
word's presentation position and the next-recalled word's presentation
position. We computed the proportions of transitions (between successively
recalled words) for each lag, normalizing for the total numbers of possible
transitions. In carrying out this analysis, we excluded all incorrect recalls
and repetitions (i.e., recalling a word that had already appeared previously in
the current recall sequence). This yielded, for each list, a 1 by
number-of-lags ($-15$ to $+15$; 30 lags in total, excluding lags of 0) array of
conditional probabilities. We averaged these probabilities across lists to
obtain a single lag-CRP for each participant. Because transitions at large
absolute lags are rare, these curves are typically displayed using range
restrictions~\citep{Kaha12}.

@@

All of the data analyzed in this manuscript, along with all of the code for
carrying out the analyses may be found at https://github.com/ContextLab/FRFR-analyses.

@@

With these differences in mind, we next compared participants' memories on
early versus late lists for two additional experimental conditions (see
\textit{Random conditions}, Fig.~\dynamicsRandom). In a \textit{reduced
(early)} condition, we held the visual features constant on early lists, but
allowed them to vary randomly on late lists. In a \textit{reduced (late)}
condition, we allowed the visual features to vary randomly on early lists, but
held them constant on late lists. Given our above findings that (a)
participants tended to exhibit stronger clustering effects on feature rich
(versus reduced) lists, and (b) participants tended to remember more words and
exhibit stronger clustering effects on early (versus late) lists, we expected
these early versus late differences to be enhanced in the reduced (early)
condition and diminished in the reduced (late) condition. However, to our
surprise, participants in \textit{neither} condition exhibited reliable early
versus late differences in accuracy (reduced (early): $t(41) = 1.499, p =
0.141$; reduced (late): $t(40) = 1.462, p = 0.152$), temporal clustering
(reduced (early): $t(41) = 0.998, p = 0.324$; reduced (late): $t(40) = 1.099, p
= 0.278$), nor feature-based clustering (reduced (early), category: $t(41) =
0.753, p = 0.456$; reduced (early), size: $t(41) = 0.721, p = 0.475$; reduced
(early), length: $t(41) = 0.493, p = 0.625$; reduced (early), first letter:
$t(41) = 0.780, p = 0.440$; reduced (late), category: $t(40) = -0.086, p =
0.932$; reduced (late), size: $t(40) = 0.746, p = 0.460$; reduced (late),
length: $t(40) = 1.476, p = 0.148$; reduced (late), first letter: $t(40) =
0.966, p = 0.340$). We hypothesized that adding or removing the variability in
the visual features was acting as a sort of ``event boundary'' between early
and late lists. In prior work, we (and others) have found that memories formed
just after event boundaries can be enhanced~\citep[e.g., due to less contextual
interference between pre- and post-boundary items;][]{MannEtal16, PettEtal16,
GoldEtal17, FlorEtal17}.

@@

If participants show different sensitivities to order manipulations, how do
their behaviors carry over to later lists? We found that participants who
showed strong feature clustering on early lists often tended to show strong
feature clustering on late lists (Fig.~\ref{fig:clustering-carryover}A; overall
across participants and conditions: $r(179) = 0.592, p < 0.001$; non-visual
feature conditions: all $r$s $\geq 0.350$, all $p$s $\leq 0.058$; color: $r(29)
= -0.071, p = 0.704$; location: $r(28) = 0.032, p = 0.868$; across conditions:
$r(4) = 0.934, p = 0.006$). Although participants tended to show weaker feature
clustering on late lists (Fig.~\ref{fig:fingerprint-trajectories}) on
\textit{average}, the associations between early and late lists for individual
participants suggests that some influence of early order manipulations may
linger on late lists. We found that participants who exhibited larger carryover
in feature clustering (i.e., continued to show strong feature clustering on
late lists) for the semantic order manipulations (but not other manipulations)
also tended to show a smaller decrease in recall on early versus late lists
(Fig.~\ref{fig:clustering-carryover}B; overall: $r(179) = 0.378, p < 0.001$;
category: $r(28) = 0.419, p = 0.021$; size: $r(28) = 0.737, p < 0.001$;
non-semantic conditions: all $r$s $\leq 0.252$, all $p$s $\geq 0.179$; across
conditions: $r(4) = 0.773, p = 0.072$) on late lists, relative to early lists.
Participants who exhibited larger carryover in feature clustering also tended
to show stronger temporal clustering on late lists (relative to early lists)
for all but the category condition (Fig.~\ref{fig:clustering-carryover}C;
overall: $r(179) = 0.434, p < 0.001$; category: $r(28) = 0.229, p = 0.223$; all
non-category conditions: all $r$s $\geq 0.448$, all $p$s $\leq 0.012$; across
conditions: $r(4) = 0.598, p = 0.210$).

@@

\subsection*{Author note}

All of the data analyzed in this manuscript, along with all of the code for
carrying out the analyses may be found at
https://github.com/ContextLab/FRFR-analyses. Code for running the non-adaptive
experimental conditions may be found at
https://github.com/Con\-text\-Lab/efficient-learning-code. Code for running the
adaptive experimental condition may be found at
https://github.com/ContextLab/adaptiveFR. We have also released an associated
Python toolbox for analyzing free recall data, which may be found at
https://cdl-quail.read\-the\-docs.io/\-en/\-latest/.  Note that this study was not
preregistered.  Some of the ideas and data presented in this manuscript
were also presented at the Annual Meeting of the Society for Neuroscience (2017).
